DCMIP2016: A Review of Nonhydrostatic Dynamical Core Design and Intercomparison of Participating Models

### 2017-04-27 DCMIP2016-Part1-Models.tex
- Added some more small modifications from James Kent

\Author[11]{David A.}{Hall}

@@

\begin{table}[p]
\caption{A standard list of symbols used throughout this paper and in the DCMIP.} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$s$ & Vertical model coordinate \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\vb{u}$ & 3D wind vector \\
$\vb{u}_h$ & Horizontal wind vector \\
$\vb{v}_h$ & Horizontal wind vector with covariant components \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence of the horizontal wind vector\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$e$ & Internal energy \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$\rho_s$ & Pseudo-density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$\theta_{il}$ & Ice-liquid potential temperature \\
%$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only three numerical methods are represented among participating models, namely finite-difference (FD), finite-volume (FV), and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\theta_{il}$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}[p]
\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Standard options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & 4th-order horizontal hyperviscosity & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, hyperviscosity & Forward-backwards \citep{LR1997QJR} / semi-implicit \\
FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
HOMME & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagornisky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
Tempest & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry, \citep{Dubos2015DYNAMICO10}.  It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature,\citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.  It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).  In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

OLAM \citep{walko2008ocean, walko2008ocean2, walko2011direct} is a deep-atmosphere, fully-compressible model that solves the equations of motion in finite-volume momentum conservation form. Acoustic modes are solved explicitly in the horizontal and implicitly in the vertical by means of time splitting. Equations are discretized with a C-grid formulation on a hexagonal Voronoi mesh with optional local mesh refinement (which introduces some pentagons and heptagons). Height is the vertical coordinate, and a Lorenz vertical grid stagger is used. A unique feature of OLAM is that grid levels are horizontal and intersect topography. This avoids a number of well-documented errors associated with terrain-following grids and also eliminates the need for evaluation of coordinate transformation terms. Topography is represented as a smooth (non-stepped) surface by means of cut cells whose surfaces and volume are reduced according to the portion of each cell that is below ground. The OLAM cut-cell formulation conserves mass and momentum.

@@

The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude.  Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.  This grid is employed by the UK Met Office \citep{davies2005new}.

@@

\begin{figure}
\centering
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=5cm]{contributions/FVM/octahedral_gauss_nodes}
 \end{subfigure}%
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=4.7cm]{contributions/FVM/octahedral_RGG_white}
 \end{subfigure}
\caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse grid with only
24 latitudes between pole and equator is used for illustration. The dual mesh resolution of the octahedral
reduced Gaussian grid is about a factor 2 finer at the poles than the equator; see \cite{smolarkiewiczetalJCP2016}.}
\label{fig:octahedralgrid}
\end{figure}

@@

\begin{figure}[]
\includegraphics[height=6cm]{gem_yinyang.png}
\caption{ Yin-Yang grid }
\label{fig:gem_yinyang}
\end{figure}

@@

ICON is based on \cite{gassmann2008} using terrain-following $z$-coordinates with prognostic density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), and vector-invariant form for horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}) and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).  The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.  The model also uses a prognostic equation for Exner pressure to simplify vertical sound wave propagation, given by
\begin{align}
\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot (\vb{u}_h \rho \theta_v) = \hat{Q},
\end{align} where $\hat{Q}$ is an appropriately formulated diabatic heat term.  The horizontal uses a Arakawa C-grid formulation on the triangular grid to prognose horizontal velocities normal to triangle edges $v_n$, making use of reconstructed tangential velocity components $v_t$.

@@

OLAM solves the deep-atmosphere, fully-compressible equations in mass- and momentum-conserving finite-volume form using equations (\ref{eq:ContinuityEquation}), (\ref{eq:PrognosticVelocityFluxForm}), and 
(\ref{eq:PotentialTemperatureFluxForm}). Prognostic variables are the 3 components of momentum, ice-liquid potential temperature $\theta_{il}$ \citep{walko2000efficient}, total density $\rho$, specific density of total water, and specific bulk density and/or bulk number concentration of various scalar quantities including liquid and ice hydrometeors, aerosols, and trace gases. For DCMIP, the latter are limited to cloud and rain specific bulk densities. Water vapor density $\rho q$ is diagnosed by subtracting bulk densities of all liquid and ice hydrometeors from the total water density, dry air density is diagnosed by subtracting total water density from total density, and pressure is diagnosed based on the equation of state and values of dry air density, water vapor density, and potential temperature $\theta$. The latter is in turn diagnosed from $\theta_{il}$ and from the latent heat required to convert any hydrometeors present to the vapor phase. Velocity components are diagnosed by dividing momentum components by total density.

@@

OLAM uses a unique temporal discretization that combines elements of the Adams-Bashforth (AB2) scheme and a Lax- Wendroff formulation for advected quantities. However, instead of extrapolating all prognostic tendencies forward to the half-future time level as in AB2, the horizontal momentum components alone (not their tendencies) are extrapolated in time at the cell boundaries where they reside. The extrapolated momentum provides the time-centered cell-to-cell total mass flux across the grid cell faces that is responsible for advective transport. Advection of all quantities, including all 3 velocity components that are diagnostically reconstructed at scalar cell centers, and advancement in time from the current to the future time level is based on the time- and space-centered Lax-Wendroff formulation. This scheme is horizontally explicit, but a trapezoidal-implicit formulation is used in the vertical for stable integration of vertically-propagating sound waves. A by-product of the implicit formulation is an implicit time-centered vertical momentum that joins the time-extrapolated horizontal momentum to form a complete set of mass fluxes for advection. Scalar advection is performed afterward and uses the same total mass fluxes as momentum advection. A time-split scheme is most often used where momentum and potential temperature are updated more frequently than other scalar fields in order to accommodate horizontally propagating sound waves.

@@

GEM differs from the approaches above by using a semi-Lagrangian advection.
Any model equations, prognostic or diagnostic, is written in the form
\begin{align}
\frac {dF}{dt}+G=0,
\label{eq:gem_temporal_O}
\end{align}
\noindent where $\frac{d}{dt}$ is the Lagrangian derivative, $F$ containing the terms subject to this operator, $G$ the remaining terms.
The semi-Lagrangian approach consists in the following space-time discretization of (\ref{eq:gem_temporal_O})
\begin{align}
\frac{F^A-F^{D}} {\Delta t}+ \left( \frac{1}{2} + \epsilon \right)G^A + \left( \frac{1}{2} - \epsilon\right )G^D = 0,
\label{eq:gem_temporal_D}
\end{align}
where $A$ stands for the arrival position at model grid point $(\vb{r}_h, \zeta, t)$
and $D$ for the departure position $(\vb{r}_h-\Delta\vb{r}_h, \zeta-\Delta \zeta, t-\Delta t)$
due to the displacements $\Delta\vb{r}_h, \Delta \zeta$
having occurred during the timestep $\Delta t$.
$G$ is averaged between these two positions with a possible slight off-centering $\epsilon$.
The displacements are themselves calculated solving, again using the Lagrangian method, the equations:
\begin{align}
\frac {d\vb{r}_h}{dt} - \vb{u}_h = 0 ; \frac {d\zeta}{dt} - \dot{\zeta} = 0,
\label{eq:gem_temporal_d_O}
\end{align}
discretized in the same way (trapezoidal method) though without off-centering:
\begin{align}
\frac{\Delta \vb{r}_h}{\Delta t} - \frac{ {\vb{u}_h}^A + {\vb{u}_h}^D }{2} = 0;
\frac{\Delta \zeta}{\Delta t} - \frac{\dot{\zeta}^A + \dot{\zeta}^D}{2} = 0.
\label{eq:gem_temporal_d_D}
\end{align}

@@

As discussed earlier, this paper represents the first in a series of papers documenting the results from the 2016 Dynamical Core Model Intercomparison Project workshop and summer school.  In this paper we have provided a description of the differences and similarities between participating models, including the choice of computational grid, horizontal staggering, vertical staggering, vertical coordinates, prognostic equations, choice of diffusion, stabilization, filters and fixers, and temporal discretization.  The literature on dynamical core development is vast, with origins that go back over half a century.  Consequently, the models discussed in this paper only represent a sample of the many dynamical cores that have been developed for general circulation modeling.  Some of the models that have not been discussed include \cite{fox1997finite, davies2005new, prusa2008eulag, nair2009computational, baba2010dynamical, donner2011dynamical, ullrich2012mcore, gassmann2013global} and \cite{doyle2014next}.

>>>>

\Author[11]{David M.}{Hall}

@@

\begin{table}[p]
\caption{A standard list of symbols used throughout this paper and in the DCMIP.} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$s$ & Vertical model coordinate \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind velocity \\
$v$ & Meridional wind velocity \\
$w$ & Vertical wind velocity \\
$\vb{u}$ & 3D wind vector \\
$\vb{u}_h$ & Horizontal wind vector \\
$\vb{v}_h$ & Horizontal wind vector with covariant components \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence of the horizontal wind vector\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$e$ & Internal energy \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$\rho_s$ & Pseudo-density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$\theta_{il}$ & Ice-liquid potential temperature \\
%$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only three numerical methods are represented among participating models, namely finite-difference (FD), finite-volume (FV), and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}[p]
\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Standard options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & 4th-order horizontal hyperviscosity & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, hyperviscosity & Forward-backwards \citep{LR1997QJR} / semi-implicit \\
FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
HOMME & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagorinsky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
Tempest & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry, \citep{Dubos2015DYNAMICO10}.  It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature \citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.  It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).  In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

OLAM \citep{walko2008ocean, walko2008ocean2, walko2011direct} is a deep-atmosphere, fully-compressible model that solves the equations of motion in finite-volume momentum conservation form. Acoustic modes are solved explicitly in the horizontal and implicitly in the vertical by means of time splitting. Equations are discretized with a C-grid formulation on a hexagonal Voronoi mesh with optional local mesh refinement (which introduces some pentagons and heptagons). Height is the vertical coordinate, and a Lorenz vertical grid staggering is used. A unique feature of OLAM is that grid levels are horizontal and intersect topography. This avoids a number of well-documented errors associated with terrain-following grids and also eliminates the need for evaluation of coordinate transformation terms. Topography is represented as a smooth (non-stepped) surface by means of cut cells whose surfaces and volume are reduced according to the portion of each cell that is below ground. The OLAM cut-cell formulation conserves mass and momentum.

@@

The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude.  Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.  This grid is employed by the UK Met Office \citep{davies2005new, wood2014inherently}.

@@

\begin{figure}
\centering
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=6cm]{contributions/FVM/octahedral_gauss_nodes}
 \end{subfigure}%
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=5.4cm]{contributions/FVM/octahedral_RGG_white}
 \end{subfigure}
\caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse grid with only
24 latitudes between pole and equator is used for illustration. The dual mesh resolution of the octahedral
reduced Gaussian grid is about a factor 2 finer at the poles than the equator; see \cite{smolarkiewiczetalJCP2016}.}
\label{fig:octahedralgrid}
\end{figure}

@@

\begin{figure}[]
\includegraphics[height=6cm]{gem_yinyang.png}
\caption{Th Yin-Yang grid, a combination of two limited-domain latitude-longitude grids assembled to provide complete coverage of the sphere.}
\label{fig:gem_yinyang}
\end{figure}

@@

ICON is based on \cite{gassmann2008} using terrain-following $z$-coordinates with prognostic density (\ref{eq:EulerianContinuityEquation}), virtual potential temperature (\ref{eq:PrognosticPseudoDensityTheta}), horizontal momentum in vector-invariant form (\ref{eq:HorizontalVelocityOrthogonal}), and vertical momentum (\ref{eq:VerticalVelocityOrthogonal}).  The specific heat capacities and ideal gas constant are approximated to be equal to their dry values $R^\ast \approx R_d$, $c^\ast_p \approx c_{pd}$ and $c^\ast_v \approx c_{vd}$.  The model also uses a prognostic equation for Exner pressure to simplify vertical sound wave propagation, given by
\begin{align}
\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot (\vb{u}_h \rho \theta_v) = \hat{Q},
\end{align} where $\hat{Q}$ is an appropriately formulated diabatic heat term.  The horizontal uses a Arakawa C-grid formulation on the triangular grid to prognose horizontal velocities normal to triangle edges $v_n$, making use of reconstructed tangential velocity components $v_t$.

@@

OLAM solves the deep-atmosphere, fully-compressible equations in mass- and momentum-conserving finite-volume form using equations (\ref{eq:EulerianContinuityEquation}), (\ref{eq:PrognosticVelocityFluxForm}), and 
(\ref{eq:PotentialTemperatureFluxForm}). Prognostic variables are the 3 components of momentum, ice-liquid potential temperature $\theta_{il}$ \citep{walko2000efficient}, total density $\rho$, specific density of total water, and specific bulk density and/or bulk number concentration of various scalar quantities including liquid and ice hydrometeors, aerosols, and trace gases. For DCMIP, the latter are limited to cloud and rain specific bulk densities. Water vapor density $\rho q$ is diagnosed by subtracting bulk densities of all liquid and ice hydrometeors from the total water density, dry air density is diagnosed by subtracting total water density from total density, and pressure is diagnosed based on the equation of state and values of dry air density, water vapor density, and potential temperature $\theta$. The latter is in turn diagnosed from $\theta_{il}$ and from the latent heat required to convert any hydrometeors present to the vapor phase. Velocity components are diagnosed by dividing momentum components by total density.

@@

OLAM uses a unique temporal discretization that combines elements of the Adams-Bashforth (AB2) scheme and a Lax- Wendroff formulation for advected quantities. However, instead of extrapolating all prognostic tendencies forward to the half-future time level as in AB2, the horizontal momentum components alone (not their tendencies) are extrapolated in time at the cell boundaries where they reside. The extrapolated momentum provides the time-centered cell-to-cell total mass flux across the grid cell faces that is responsible for advective transport. Advection of all quantities, including all 3 velocity components that are diagnostically reconstructed at scalar cell centers, and advancement in time from the current to the future time level is based on the time- and space-centered Lax-Wendroff formulation.   This scheme is horizontally explicit, but a trapezoidal- implicit formulation is used in the vertical for stable integration of vertically-propagating sound waves. A by-product of the implicit formulation is an implicit time-centered vertical momentum that joins the time-extrapolated horizontal momentum to form a complete set of mass fluxes for advection. The vertical momentum equation is solved first so that the time-centered vertical momentum is available for computing transport of horizontal momentum and all scalar quantities. A time-split scheme is most often used where momentum and potential temperature are updated more frequently than other scalar fields in order to accommodate horizontally propagating sound waves.

%This scheme is horizontally explicit, but a trapezoidal-implicit formulation is used in the vertical for stable integration of vertically-propagating sound waves. A by-product of the implicit formulation is an implicit time-centered vertical momentum that joins the time-extrapolated horizontal momentum to form a complete set of mass fluxes for advection. Scalar advection is performed afterward and uses the same total mass fluxes as momentum advection. A time-split scheme is most often used where momentum and potential temperature are updated more frequently than other scalar fields in order to accommodate horizontally propagating sound waves.

@@

GEM differs from the approaches above by using a semi-Lagrangian advection.
Any model equations, prognostic or diagnostic, are written in the form
\begin{align}
\frac {dF}{dt}+G=0,
\label{eq:gem_temporal_O}
\end{align}
\noindent where $\frac{d}{dt}$ is the Lagrangian derivative, $F$ containing the terms subject to this operator, $G$ the remaining terms.
The semi-Lagrangian approach consists in the following space-time discretization of (\ref{eq:gem_temporal_O})
\begin{align}
\frac{F^A-F^{D}} {\Delta t}+ \left( \frac{1}{2} + \epsilon \right)G^A + \left( \frac{1}{2} - \epsilon\right )G^D = 0,
\label{eq:gem_temporal_D}
\end{align}
where $A$ stands for the arrival position at model grid point $(\vb{r}_h, \zeta, t)$
and $D$ for the departure position $(\vb{r}_h-\Delta\vb{r}_h, \zeta-\Delta \zeta, t-\Delta t)$
due to the displacements $\Delta\vb{r}_h, \Delta \zeta$
having occurred during the timestep $\Delta t$.
$G$ is averaged between these two positions with a possible slight off-centering $\epsilon$.
The displacements are themselves calculated solving, again using the Lagrangian method, the equations:
\begin{align}
\frac {d\vb{r}_h}{dt} - \vb{u}_h = 0 ; \frac {d\zeta}{dt} - \dot{\zeta} = 0,
\label{eq:gem_temporal_d_O}
\end{align}
discretized in the same way (trapezoidal method) though without off-centering:
\begin{align}
\frac{\Delta \vb{r}_h}{\Delta t} - \frac{ {\vb{u}_h}^A + {\vb{u}_h}^D }{2} = 0;
\frac{\Delta \zeta}{\Delta t} - \frac{\dot{\zeta}^A + \dot{\zeta}^D}{2} = 0.
\label{eq:gem_temporal_d_D}
\end{align}

@@

As discussed earlier, this paper represents the first in a series of papers documenting the results from the 2016 Dynamical Core Model Intercomparison Project workshop and summer school.  In this paper we have provided a description of the differences and similarities between participating models, including the choice of computational grid, horizontal staggering, vertical staggering, vertical coordinates, prognostic equations, choice of diffusion, stabilization, filters and fixers, and temporal discretization.  The literature on dynamical core development is vast, with origins that go back over half a century.  Consequently, the models discussed in this paper only represent a sample of the many dynamical cores that have been developed for general circulation modeling.  Some of the models that have not been discussed include \cite{fox1997finite, prusa2008eulag, nair2009computational, baba2010dynamical, donner2011dynamical, ullrich2012mcore, gassmann2013global, wood2014inherently} and \cite{doyle2014next}.
